probKanren: A Simple Probabilistic extension for microKanren

### 2021-08-23 paper/main.tex
Add updated code to repo

%%
%% Minted listings support 
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
\usepackage{minted}
%% auto break lines
\setminted{breaklines=true}

@@

%%
%% This command is for the conference information
\conference{Woodstock'21: Symposium on the irreproducible science,
  June 07--11, 2021, Woodstock, NY}

%%
%% The "title" command
\title{probKanren: A Simple Probabilistic extension for miniKanren}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
\author[1]{Robert Zinkov}[%
%orcid=0000-0002-0877-7063,
email=zinkov@robots.ox.ac.uk,
url=https://zinkov.com/,
]
\address[1]{Peoples' Friendship University of Russia (RUDN University),
  6 Miklukho-Maklaya St, Moscow, 117198, Russian Federation}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Probabilistic programming is sometimes described as a generalisation
  of logic programming where instead of just returning a set of facts
  that answer a given query, which return a probability distribution over
  facts that answer a given query. But many contemporary probabilistic
  logic programming languages are not implemented as simple extensions
  of existing logic programming languages but instead involve their
  own unique implementations. Here we introduce probKanren, a simple
  extension to miniKanren that transforms
\end{abstract}

@@

@@

@@

\subsection{Contributions}

@@

There is a rich history of extending logic programming formalisms to
support probabilistic inference. Early systems like
PRISM\cite{sato1997prism} and ProbLog\cite{de2007problog} allowed
associating discrete distributions with facts. Later
work\cite{gutmann2010extending} introduces distribution clauses so
that some continuous distributions. Other work extended these methods
further while focusing on efficient exact inference algorithms like
model weight integration\cite{islam2012inference,
  belle2015probabilistic}. Our work is most similar to
\cite{gutmann2011magic} except that while they combine their forward
reasoning with an importance sampler we use a particle cascade instead
which can be more sample efficient.

@@

\subsection{miniKanren}

\textsc{miniKanren}\cite{byrd2017unified, daniel2018reasoned} is a
pure logic programming language embedded in Scheme. The language
consists of a set of terms, functions for constructing, a set of goal
primitives, and two run functions to answer queries in the
language. The goal primitives consist of a \texttt{fresh} form for
introducing logic variables, a unification primitive, a conjunction
combinator, and a disjunction combinator.

Define terms, goals, and substitutions here

@@

\section{Grammar and Definitions}

Define distributional clauses here

@@

We propose to extend \textsc{miniKanren} by augmenting each of the
search streams with a set of particles. These particles represent the
empirical distribution of that stream. Each particle has associated
with it a substitution of all the logic and random variables as well
as a weight that is proportional to the likelihood of the substitution.

An initial set of particles is created from the probabilistic program
when it is first run. As disjunctions like \texttt{conde} are encountered,
we split evenly the number of particles allocated to each stream. Whenever
we encounter a unification primitive, we run a resampling step. This
helps to prune low-weight particles and replicate high weight ones.

@@

Firstly, the arguments of distribution clauses must be
ground. Secondly, a random variable can not unify with any arithmetic
expression

@@

One of the experiments should be something from the probabilistic logic literature.

Friends who Smoke is a probabilistic logic program which models the social nature of
who smokes cigarettes. The model predicts that people who are friends with people
who smoke are more likely to smoke.

We could include experiment from Noah Goodman's dippl work like a semantic parsing
example\\ \url{http://dippl.org/examples/semanticparsing.html}

We could include program synthesis experiment, where probabilities
allow us to specify a soft preference for using certain language
primitives in a similar spirit to
\textsc{rKanren}\cite{swords2013rkanren} or neural-guided search
\cite{zhang2018neural}.

@@

We made a cool and simple to implement extension to
\textsc{miniKanren} that let's us support probabilistic inference on
both discrete and continuous distributions.

>>>>

%%
%% Minted listings support 
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
\usepackage{minted}
\usepackage{syntax}
\shortverb{\|}
%% auto break lines
\setminted{breaklines=true}

@@

%%
%% This command is for the conference information
\conference{PLP '21: The 8th Workshop on Probabilistic Logic Programming,
  September 20--27, 2021, Porto, Portugal}

%%
%% The "title" command
\title{probKanren: A Simple Probabilistic extension for microKanren}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
\author[1]{Robert Zinkov}[%
%orcid=0000-0002-0877-7063,
email=zinkov@robots.ox.ac.uk,
url=https://zinkov.com/,
]
\address[1]{Dept. of Engineering Science, University of Oxford,
  25 Banbury Rd, Oxford, UK}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Probabilistic programming is sometimes described as a generalisation
  of logic programming where instead of just returning a set of
  answers to a given query, we also return a probability distribution
  over those answers. But many contemporary probabilistic logic
  programming languages are not implemented as simple extensions of
  existing logic programming languages but instead involve their own
  unique implementations. Here we introduce probKanren, a simple
  extension to microKanren that transforms it into a probabilistic
  programming language without needing to make any modifications to
  the underlying search.
\end{abstract}

@@

But the conceptual simplicity of this sort of generalisation is not reflected
in the complexity of many existing probabilistic logic programming systems.
They often involve making implementing sophisticated algorithms and the underlying
systems are not just implemented on top of existing logic programming systems.

We believe a conceptually simple extension deserves a conceptually
simple implementation to go along with it. We thus contribute a simple
way to extend microkanren a small logic programming DSL such that it
becomes a probabilistic programming language.

@@

\begin{minted}{Scheme}
  (run 1000 (q)
    (beta 1 1 q)
    (flip q #t))
\end{minted}

@@

@@

There is a rich history of extending logic programming formalisms to
support probabilistic inference. Early systems like
PRISM\cite{sato1997prism} and ProbLog\cite{de2007problog} allowed
associating discrete distributions with facts. Early versions of
ProbLog were also built on top of Prolog matching one of the goals of
our work. Later work\cite{gutmann2010extending} introduces
distribution clauses so that some continuous distributions. Other work
extended these methods further while focusing on efficient exact
inference algorithms like model weight
integration\cite{islam2012inference, belle2015probabilistic}. Our work
is most similar to \cite{gutmann2011magic} except that while they
combine their forward reasoning with an importance sampler we use a
particle cascade instead which can be more sample efficient.

@@

\subsection{microKanren}

\textsc{microKanren}\cite{10.1145/2989225.2989230, daniel2018reasoned}
is a pure logic programming language embedded in Scheme. The language
consists of a set of terms, a set of goal primitives, and two run
functions to answer queries in the language. The goal primitives
consist of a \texttt{fresh} form for introducing logic variables, a
unification primitive, a conjunction combinator, a disjunction
combinator, and ways to define and use relations.

\subsection{Grammar and Definitions}

\begin{figure}
\begin{grammar}
<prog> ::= (run <number> (<id>) <goal-expr>)

<goal-expr> ::= (disj <goal-expr> <goal-expr>) \alt
	        (conj <goal-expr> <goal-expr>) \alt
                (fresh (<id>) <goal-expr>) \alt
		(== <term-expr> <term-expr>) \alt
		(letrec-rel ((<id> (<id> ...) <goal-expr>) ...) \\
		\hspace{\grammarindent} <goal-expr>) \alt
		(call-rel <lexical-var-ref> <term-expr> ...) \alt
		(prim-rel-call <lexical-var-ref> <term-expr> ...) \alt
		(delay <goal-expr>)

<term-expr> ::= (quote <datum>) \alt
                <lexical-var-ref> \alt
                (cons <term-expr> <term-expr>)

<term> ::= <number> \alt
           \#f | \#t \alt
	   <symbol> \alt
	   (<term> . <term>) \alt
	   <logic-var>

\end{grammar}
\end{figure}

To create probkanren we extend this grammar with distribution clauses
such as \texttt{normal} and \texttt{flip}. These are just another type
of goal expression. We do not need to add a notion of probabilistic
variables as they can just be seen as logic variables constrained in
a particular way.

@@

@@

We propose to extend \textsc{microKanren} by augmenting each of the
search streams with a set of particles. These particles represent the
empirical distribution of that stream. Each particle has associated
with it a substitution of all the logic and random variables as well
as a weight that is proportional to the likelihood of the substitution.

An initial set of particles is created from the probabilistic program
when it is first run. As disjunctions (\texttt{disj}) are encountered,
we split evenly the number of particles allocated to each stream. Whenever
we encounter a unification primitive, we run a resampling step. This
helps to prune low-weight particles and replicate high weight ones.

@@

Firstly, the arguments of distribution clauses must be
ground. Secondly, a random variable cannot unify with any arithmetic
expression.

This extension does not modify the search and streams are managed
exactly as in microkanren. An additional advantage of this is thanks
to the microkanren search being complete, if we generate enough
particles we are guaranteed to recover the true posterior as all paths
of the search space will eventually be explored.

@@

We validate that probKanren is at least as expressive as other probabilistic
logic programming languages by implementing the Friends who Smoke model.

Friends who Smoke is a probabilistic logic program which models the
social nature of who smokes cigarettes. The model predicts that people
who are friends with people who smoke are more likely to smoke. We
replicate the example on
\url{https://dtai.cs.kuleuven.be/problog/tutorial/basic/05_smokers.html}
using 2000 particles and get an empirical distribution that seems to
match up with the discrete distribution returned from ProbLog.

%% Compare to birch or Anglican on some top examples as well

%% We could include experiment from Noah Goodman's dippl work like a semantic parsing
%% example\\ \url{http://dippl.org/examples/semanticparsing.html}

%% We could include program synthesis experiment, where probabilities
%% allow us to specify a soft preference for using certain language
%% primitives in a similar spirit to
%% \textsc{rKanren}\cite{swords2013rkanren} or neural-guided search
%% \cite{zhang2018neural}.

@@

We made a simple to implement extension to \textsc{microKanren} that
let's us support probabilistic inference on both discrete and
continuous distributions. The approach does not require modifying the
underlying search algorithm or touch any of the backtracking code and
comes with a theoretical guarantee that if the underlying search is
complete then the probabilistic extension will require the true
posterior given enough particles.
